---
title: "Lab 4: Networks"
subtitle: "BMI 206"
author: "Isobel Beasley"
date: "2024-11-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 1. After exploring the Manhattan plots, and qq-plots from each GWAS, what can you tell about the power of each study?

```{r q1_run, eval = F}
source(here::here("networks/Manhattan_plot.R"))
source(here::here("networks/qqplot.R"))
```

```{r q1_view_1, echo=FALSE, results='asis'}
cat('<iframe src="Manhattan_plot.pdf" width="950px" height="525px"></iframe>')
```

```{r q1_view_2, echo=FALSE, results='asis'}
cat('<iframe src="qqplot.pdf" width="500px" height="550px"></iframe>')
```

<br>
<br>

## Question 2. Using Cytoscape, analyze the PPI and describe its main network.

<br>
<br>

## Question 3. Using Cytoscape, find the first order networks (p<0.05) for each GWAS.


<br>
<br>

## Question 4. Source “Pathway_permutation.r”. Are the first order networks from both GWAS more connected than expected? What does this mean?

```{r q4_run, results='hide', eval = F}
#source(here::here("networks/Pathway_permutations.R"))
system(paste("RScript",
             here::here("networks/Pathway_permutation.R"), 
             "-p networks/parent_PPI.sif",
             "-o networks/q4_pathway_permutation.pdf"
             )
       )
```

```{r q4_view, echo=FALSE, results='asis'}
cat('<iframe src="q4_pathway_permutation.pdf" width="900px" height="950px"></iframe>')
```

<br>
<br>

## Question 5. Run BINGO App on all nodes from largest connected component. What biological processes emerge from the first order networks?


<br>
<br>

## Question 6. Map and color known MS and HT genes onto their respective first order nets. Interpret results.

<br>
<br>

## Question 7. Repeat steps 3 and 4 with directed protein network from PNAS paper. 

```{r q7_run, results='hide', eval = F}
#source(here::here("networks/Pathway_permutations.R"))
system(paste("RScript",
             here::here("networks/Pathway_permutation.R"), 
             "-p networks/Directed_PPI.sif",
             "-o networks/q7_pathway_permutation.pdf"
             )
       )
```

```{r q7_view, echo=FALSE, results='asis'}
cat('<iframe src="q4_pathway_permutation.pdf" width="900px" height = "950px"></iframe>')
```


<br>
<br>

## Question 8. Color nodes by controllability category (dispensable, indispensable, neutral).


<br>
<br>

## Question 9. Repeat step 6. Are MS-associated genes more enriched in any controllability category? Interpret. 


<br>
<br>


