---
title: 'Lab 1: Linear models for quantitative genetics'
subtitle: "BMI 206"
author: "Isobel Beasley"
date: "Due: 18th October 20224"
output: html_document
---
<br>
<br>

## PART1: Analyzing provided genotype and phenotype data.

### __Prepare the data.__
Read in the genotype and phenotype matrices. 
```{r}
genos = as.matrix(read.table(here::here("gwas/genos.txt")))
phenos = as.matrix(read.table(here::here("gwas/phenos.txt")))
```
<br>

### Make a histogram of the phenotypes. Do they look normally distributed?

```{r}
hist(phenos)
mean(phenos)
median(phenos)
```
<br>

Yes, the phenotypes appear roughly normally distributed: single mode / mean, distributed roughly symmetrically about this mean (with perhaps the exception a few left skewed values, but the mean and median values are still quite similar). Additionally, the tails of this distribution do not appear particularly 'fat', meaning the distribution is likely better approximated by a normal distribution than t-distribution. 

### How are the genotypes encoded?
```{r}
table(genos)
```
The genotypes are encoded such that the heterozygous genotype is 1, the homogyzous genotypes are encoded as 2 or 0, depending on which allele is consider the reference at each locus.

<br>

### How many individuals are there in the dataset and how many SNPs? (Save them in `N` and `M`, respectively.)
```{r, eval=TRUE}
dim(genos)
dim(phenos)
N = length(phenos)
N # number of individuals
M = ncol(genos)
M # number of SNPs
```

<br>

### __Compute the *minor* allele frequency for every SNP. Check MAFs are <0.5.__

```{r, eval=TRUE}
MAFs = array(0,M)

for(i in 1:M) {
      MAFs[i] = sum(genos[,i]) / (2*N)
      
      # checking that MAF<0.5
      # if major allele freq was calculated, then 
      # take minor allele freq as 1-allele freq
      # if(MAFs[i]>0.5){
      #   MAFs[i] = 1-MAFs[i]
      # }
      
      MAFs[i] = min(MAFs[i], 1-MAFs[i])
      
}

# out of interest, this was the fastest approach I tried
# MAFs = colSums(genos) / (2*N)
# MAFs = ifelse(MAFs > 0.5, 1-MAFs, MAFs)

length(MAFs) # check that it was calculated for the right number of snps
MAFs[1:10] # how does the MAF look for the first 10 snps 
max(MAFs)
```
<br>

### __Run a GWAS under an additive model and save the p-values, z-scores, and effect sizes.__
```{r, eval=TRUE}

pvalues = array(0,M)
zscores = array(0,M)
betas = array(0,M)

for(i in 1:M) {
	g = genos[,i]
	# under an additive model means, without considering possible dominance / recessives
	res = summary(lm(phenos~g))
	zscores[i] = res$coefficients[2,"t value"]
	pvalues[i] = res$coefficients[2,"Pr(>|t|)"]
	betas[i] = res$coefficients[2,"Estimate"]
}
```
<br>

### Summarize the effect sizes.

```{r, eval=TRUE}
summary(betas)
hist(betas, breaks = 20, main = "Distribution of Estimated Effect sizes")

```
Most effect sizes are very close to zero (very small). Positve and ne
Effect sizes are roughly equally likely to be positive, as the are negative (when all snps are considered). 
<br>

### Are there any significantly associated SNPs? If so, which SNPs are they?
```{r,eval=TRUE}
sig_thresh = 0.05 / M #bonferroni corrected significance test 
assoc = which(pvalues<sig_thresh)
assoc # the first 10 snps are the significantly associated sns
```
<br>

### How big are their effect sizes? How significant are they? 
```{r, eval=TRUE}
betas[assoc] # this is how big their effect sizes are
zscores[assoc] # zscores tell you about significance - but more importantly
pvalues[assoc] # the p-values tell you how significant they are ... pretty significant, which makes sense given we use the more conservative bonferroni correction 
```
<br>

### Draw a QQ plot for log10(p) values.

```{r, eval=TRUE, fig.height = 9}
obsLogPvs = sort(-log10(pvalues))
expLogPvs = sort(-log10(seq(1/M,1,1/M)))
plot(expLogPvs,obsLogPvs,main='QQ plot')
abline( a=0, b=1 )
#label the significant SNPs red 
points(expLogPvs[(M-length(assoc)):M],obsLogPvs[(M-length(assoc)):M],col="red")
```
<br>

### Is there inflation? Use the chi-square statistics to check.
```{r, eval=TRUE}
chis = zscores^2
lambdaGC = median(chis)/0.454 # why .454?
# .454 =  qchisq(0.5, 1), the median of chisquared distribution with 1 degree of freedom (i.e. the median of null distribution)
lambdaGC
# this value is close to 1, suggestive of no inflation 
```
<br>

### Plot the phenotype predictions for the most significant SNP.
```{r, eval=TRUE}
topSNP = genos[,order(pvalues)[1]]
plot(topSNP,phenos)
abline(lm(phenos~topSNP)$coeff,col="red")
```
<br>

### __Build a linear predictor of the phenotype using the associated SNPs.__
```{r, eval=TRUE}
ypred = array(0,N)
# for each individual, predict their phenotypic score as the sum of their genotypes
# times the respective effect size calculated for that snp
for(i in 1:N) {
      ypred[i] = genos[i,assoc] %*% betas[assoc]
}
plot(ypred,phenos)
```
<br>

### What is the correlation between the predicted phenotype and the true phenotype?

```{r, eval=TRUE}
cor(ypred,phenos)
```
<br>

### __BONUS: Test each of the associated SNPs for non-linearity.__
### (stuck) Checking how test for non-linearity
```{r, eval=TRUE}

hp = array(0,length(assoc))

for (i in 1:length(assoc)) {
  g = genos[,assoc[i]]
  h = g
  h[h==2]=0 #stuck (? not 100% sure of)
  #Hint: can use anova(lm(?),lm(?)) or summary(lm(?))
  # ??? really not sure about this, testing against just intercept
  hp[i] <- anova( lm(phenos~h), lm(phenos~1) )$Pr[2] #skip multiple test correction for now
}
hp 


for (i in 1:length(assoc)) {
  g = genos[,assoc[i]]
  h = g
  h[h==2]=0
  #Hint: can use anova(lm(?),lm(?)) or summary(lm(?))
  # ??? really not sure about this, testing against just intercept
  hp[i] <- anova( lm(phenos~h), lm(phenos~1) )$Pr[2] #skip multiple test correction for now
}
```
<br>

### BONUS: Visualize a linear SNP and a non-linear SNP.
```{r, eval=TRUE}

par(mfrow=c(1,2) )

# get linear snp
lin_snp = assoc[which(hp==max(hp))]

# plot linear snp
plot(genos[,lin_snp], phenos)
points( c(0,1,2), tapply(phenos,genos[,lin_snp], mean ), col=2, pch=16, cex=3 )
lines( c(0,1,2), tapply(phenos,genos[,lin_snp], mean ), col=2, lwd=2  )

# non-linear snp

# get non-linear snp
non_lin_snp = assoc[which(hp==min(hp))]

# plot non-linear snp
plot( genos[,non_lin_snp], phenos )
points( c(0,1,2), tapply(phenos, genos[,non_lin_snp], mean ), col=2, pch=16, cex=3 )
lines( c(0,1,2), tapply(phenos, genos[,non_lin_snp], mean ), col=2, lwd=2  )

```
<br>

### __Repeat the GWAS to test for recessive rather than additive genetic effects.__
```{r, eval=TRUE}
genos2 = genos
genos2[genos<2]=1 # ?? this sets 0,1 as 1 - but i'd be inclined to set 0,1 as 0, and 2 as 1. 
pvalues2 = array(0,M)
zscores2 = array(0,M)
betas2 = array(0,M)
for(i in 1:M) {
  g = genos2[,i]
  res = summary(lm(phenos~g))
  zscores2[i] = res$coefficients[2,"t value"]
  pvalues2[i] = res$coefficients[2,"Pr(>|t|)"]
  betas2[i] = res$coefficients[2,"Estimate"]
}

```
<br>

### __Are the same SNPs significant or not?__
```{r, eval=TRUE}

assoc2 = which(pvalues2<sig_thresh)
assoc2

setdiff(assoc2, assoc)
setdiff(assoc, assoc2)

```
<br>

### __How did the effect sizes change?__
```{r, eval=TRUE}
plot(betas,betas2)
```
<br>

## PART2: Simulating genotypes with LD.

### __Establish some important simulation parameters.__
```{r, eval=TRUE}
N = 1000 #number of individuals
M = 30   #number of non-causal SNPs
gs = matrix(0,nrow=N,ncol=M)
```
<br>

### __Simulate a GWAS data set.__
#### First, simulate the causal variant.
```{r, eval=TRUE}
set.seed = (42) #set random seed so we all get the same numbers
MAF = 0.5
gC = rbinom(N,1,MAF) #causal variant
```
<br>

#### Then, simulate the phenotypes given the causal variant.
```{r, eval=TRUE}
beta = 0.3 #association of causal variant
pheno = gC*beta + rnorm(N) 
```
<br>

#### Generate 10 SNPS in tight LD with the causal SNP.

#### (stuck on prove they have the right LD theoretically)
```{r, eval=TRUE}
rho = 0.9
for(i in 1:10) {
  idx = rbinom(N,1,rho)
  gs[,i]=gC*idx+rbinom(N,1,MAF)*(1-idx)
  # test they have the right LD empirically
  cat( 'Observed LD = ', cor( gs[,i], gC ), '\n' )
  # Bonus: prove they have the right LD theoretically
}
```
<br>

#### Do the same for 10 moderate LD partners (rho=0.6).
```{r,eval=TRUE}
rho = 0.6
for(i in 1:10) {
  idx = rbinom(N,1,rho)
  gs[,i+10]=gC*idx+rbinom(N,1,MAF)*(1-idx)
  # test they have the right LD empirically
  cat( 'Observed LD = ', cor( gs[,i+10], gC ), '\n' )
  # Bonus: prove they have the right LD theoretically
}
```
<br>

Do the same for 10 independent SNPs (rho=0).
```{r,eval=TRUE}
rho = 0
for(i in 1:10) {
  idx = rbinom(N,1,rho)
  gs[,i+20]=gC*idx+rbinom(N,1,MAF)*(1-idx)
  # test they have the right LD empirically
  cat( 'Observed LD = ', cor( gs[,i+20], gC ), '\n' )
}
```

### __Run GWAS on the causal variant. Then run GWAS on the other variants. Keep track of the zscores only.__
```{r,eval=TRUE}
zsC = summary(lm(pheno~gC))$coef[2,3]
zs = sapply( 1:M, function(i) summary(lm(pheno~gs[,i]))$coef[2,3] )
```
<br>

### (stuck on ... not entirely sure correct) Visualize the relationship between the mean z-scores at the tag SNPs and the z-score at the causal SNP.
```{r,eval=TRUE}
par( mfrow=c(2,2) )
breaks = hist(c(0,zsC,zs),plot=F)$breaks
hist(zs[1:10],breaks=breaks, col=1, main='LD partners')
abline(v=zsC)
hist(zs[11:20],breaks=breaks, col=2, main='Low-LD partner SNPs')
abline(v=zsC)
hist(zs[21:30],breaks=breaks, col=3, main='Independent SNPs')
abline(v=zsC)
```
<br>

## Quite stuck here (Date: 14/10)

### __BONUS: Perform LD score regression. First, calculate the LD scores. There should be M+1 of them.__
```{r, eval=TRUE}
# ? LD scores are: the sum the squared Pearson correlation coefficients (r²) of a given SNP with all other SNPs 
all_gs = cbind(gs, gC)
cor_mat = apply(all_gs, MARGIN = 2, function(geno) cor(geno, gs))
colSums(cor_mat^2 - 1)

ldscores = colSums(cor_mat^2 - 1)
ldscores 
length(ldscores) # M + 1 = 31 - correct number of ldscores
```
<br>

### BONUS: Visualize LD score regression.
```{r,eval=FALSE}
# expected [chisq | ] = N h ^ 2 lj + N a + 1
# https://www.nature.com/articles/ng.3211
#chis = c( ?, ? )^2
chis = c( 1, 2 )^2
plot(ldscores[1:2], chis, ylab=expression(chi^2) )
#test for inflation
lambdaGC = median(chis)/0.454
lambdaGC
```
<br>

### BONUS: Estimate heritability.
```{r,eval=FALSE}
summary( lm( ? )$coef[2,1] * M/N
```
<br>

### BONUS: What is the true heritability?
```{r, eval=FALSE}
# narrow sence: phenotypic variation / genotypic variation
var(?) / var(sum())
```









