---
title: 'Lab 1: Linear models for quantitative genetics'
subtitle: "BMI 206"
author: "Isobel Beasley"
date: "18th October 20224"
output: html_document
---
<br>
<br>

## PART1: Analyzing provided genotype and phenotype data.

### __Prepare the data.__
Read in the genotype and phenotype matrices. 
```{r}
genos = as.matrix(read.table(here::here("gwas/genos.txt")))
phenos = as.matrix(read.table(here::here("gwas/phenos.txt")))
```
<br>

### Make a histogram of the phenotypes. Do they look normally distributed?

```{r}
hist(phenos)
```
<br>

### How are the genotypes encoded?
```{r}
table(genos)
```
<br>

### How many individuals are there in the dataset and how many SNPs? (Save them in `N` and `M`, respectively.)
```{r, eval=TRUE}
dim(genos)
dim(phenos)
N = length(phenos)
N
M = ncol(genos)
M
```

<br>

### __Compute the *minor* allele frequency for every SNP. Check MAFs are <0.5.__
```{r, eval=TRUE}
MAFs = array(0,M)

for(i in 1:M) {
      MAFs[i] = sum(genos[,i]) / (2*N)
      
      # if major allele freq was calculated, then 
      # calculate minor allele freq as 1-allele freq
      if(MAFs[i]>0.5){
        MAFs[i] = 1-MAFs[i]
      }
}
MAFs[1:10]
max(MAFs)
```
<br>

### __Run a GWAS under an additive model and save the p-values, z-scores, and effect sizes.__
```{r, eval=TRUE}
pvalues = array(0,M)
zscores = array(0,M)
betas = array(0,M)
for(i in 1:M) {
	g = genos[,i]
	res = summary(lm(phenos~g))
	zscores[i] = res$coefficients[2,"t value"]
	pvalues[i] = res$coefficients[2,"Pr(>|t|)"]
	betas[i] = res$coefficients[2,"Estimate"]
}
```
<br>

### Summarize the effect sizes.

```{r, eval=TRUE}
summary(betas)
hist(betas)
```
<br>

Are there any significantly associated SNPs? If so, which SNPs are they?
```{r,eval=TRUE}
sig_thresh = 0.05 / M #bonferroni corrected significance test 
assoc = which(pvalues<sig_thresh)
assoc
```
<br>

### How big are their effect sizes? How significant are they? 
```{r, eval=TRUE}
betas[assoc]
zscores[assoc]
pvalues[assoc]
```
<br>

### Draw a QQ plot for log10(p) values.

```{r, eval=TRUE}
obsLogPvs = sort(-log10(pvalues))
expLogPvs = sort(-log10(seq(1/M,1,1/M)))
plot(expLogPvs,obsLogPvs,main='QQ plot')
abline( a=0, b=1 )
#label the significant SNPs red 
points(expLogPvs[(M-length(assoc)):M],obsLogPvs[(M-length(assoc)):M],col="red")
```
<br>

Is there inflation? Use the chi-square statistics to check.
```{r, eval=TRUE}
chis = zscores^2
lambdaGC = median(chis)/0.454 # why .454?
# .454 =  qchisq(0.5, 1), the median of chisquared distribution with 1 degree of freedom (i.e. the median of null distribution)
lambdaGC
# this value is close to 1, suggestive of no inflation 
```
<br>

Plot the phenotype predictions for the most significant SNP.
```{r, eval=TRUE}
topSNP = genos[,order(pvalues)[1]]
plot(topSNP,phenos)
abline(lm(phenos~topSNP)$coeff,col="red")
```
<br>

### __Build a linear predictor of the phenotype using the associated SNPs.__
```{r, eval=TRUE}
ypred = array(0,N)
for(i in 1:N) {
      ypred[i] = genos[i,assoc] %*% betas[assoc]
}
plot(ypred,phenos)
```
<br>

### What is the correlation between the predicted phenotype and the true phenotype?

```{r, eval=TRUE}
cor(ypred,phenos)
```
<br>

### __BONUS: Test each of the associated SNPs for non-linearity.__
```{r, eval=TRUE}
hp = array(0,length(assoc))
for (i in 1:length(assoc)) {
  g = genos[,assoc[i]]
  h = g
  h[h==2]=0
  #Hint: can use anova(lm(?),lm(?)) or summary(lm(?))
  # ??? really not sure about this, testing against just intercept
  hp[i] <- anova( lm(phenos~h), lm(phenos~1) )$Pr[2] #skip multiple test correction for now
}
hp 
```
<br>

### BONUS: Visualize a linear SNP and a non-linear SNP.
```{r, eval=TRUE}

par(mfrow=c(1,2) )

# get linear snp
lin_snp = assoc[which(hp==max(hp))]

# plot linear snp
plot(genos[,lin_snp], phenos)
points( c(0,1,2), tapply(phenos,genos[,lin_snp], mean ), col=2, pch=16, cex=3 )
lines( c(0,1,2), tapply(phenos,genos[,lin_snp], mean ), col=2, lwd=2  )

# non-linear snp

# get non-linear snp
non_lin_snp = assoc[which(hp==min(hp))]

# plot non-linear snp
plot( genos[,non_lin_snp], phenos )
points( c(0,1,2), tapply(phenos, genos[,non_lin_snp], mean ), col=2, pch=16, cex=3 )
lines( c(0,1,2), tapply(phenos, genos[,non_lin_snp], mean ), col=2, lwd=2  )

```
<br>

### __Repeat the GWAS to test for recessive rather than additive genetic effects.__
```{r, eval=TRUE}
genos2 = genos
genos2[genos<2]=1 # ?? this sets 0,1 as 1 - but i'd be inclined to set 0,1 as 0, and 2 as 1. 
pvalues2 = array(0,M)
zscores2 = array(0,M)
betas2 = array(0,M)
for(i in 1:M) {
  g = genos2[,i]
  res = summary(lm(phenos~g))
  zscores2[i] = res$coefficients[2,"t value"]
  pvalues2[i] = res$coefficients[2,"Pr(>|t|)"]
  betas2[i] = res$coefficients[2,"Estimate"]
}

```
<br>

### __Are the same SNPs significant or not?__
```{r, eval=TRUE}

assoc2 = which(pvalues2<sig_thresh)
assoc2

setdiff(assoc2, assoc)
setdiff(assoc, assoc2)

```
<br>

### __How did the effect sizes change?__
```{r, eval=TRUE}
plot(betas,betas2)
```
<br>

## PART2: Simulating genotypes with LD.

### __Establish some important simulation parameters.__
```{r}
N = 1000 #number of individuals
M = 30   #number of non-causal SNPs
gs = matrix(0,nrow=N,ncol=M)
```
<br>

### __Simulate a GWAS data set.__
#### First, simulate the causal variant.
```{r}
set.seed = (42) #set random seed so we all get the same numbers
MAF = 0.5
gC = rbinom(N,1,MAF) #causal variant
```
<br>

#### Then, simulate the phenotypes given the causal variant.
```{r}
beta = 0.3 #association of causal variant
pheno = gC*beta + rnorm(N) 
```
<br>

### Generate 10 SNPS in tight LD with the causal SNP.
```{r}
rho = 0.9
for(i in 1:10) {
  idx = rbinom(N,1,rho)
  gs[,i]=gC*idx+rbinom(N,1,MAF)*(1-idx)
  # test they have the right LD empirically
  cat( 'Observed LD = ', cor( gs[,i], gC ), '\n' )
  # Bonus: prove they have the right LD theoretically
}
```
<br>

#### Do the same for 10 moderate LD partners (rho=0.6).
```{r,eval=TRUE}
rho = 0.6
for(i in 1:10) {
  idx = rbinom(N,1,rho)
  gs[,i+10]=gC*idx+rbinom(N,1,MAF)*(1-idx)
  # test they have the right LD empirically
  cat( 'Observed LD = ', cor( gs[,i+10], gC ), '\n' )
  # Bonus: prove they have the right LD theoretically
}
```
<br>

Do the same for 10 independent SNPs (rho=0).
```{r,eval=TRUE}
rho = 0
for(i in 1:10) {
  idx = rbinom(N,1,rho)
  gs[,i+20]=gC*idx+rbinom(N,1,MAF)*(1-idx)
  # test they have the right LD empirically
  cat( 'Observed LD = ', cor( gs[,i+20], gC ), '\n' )
}
```

### __Run GWAS on the causal variant. Then run GWAS on the other variants. Keep track of the zscores only.__
```{r,eval=TRUE}
zsC = summary(lm(pheno~gC))$coef[2,3]
zs = sapply( 1:M, function(i) summary(lm(pheno~gs[,i]))$coef[2,3] )
```
<br>

### Visualize the relationship between the mean z-scores at the tag SNPs and the z-score at the causal SNP.
```{r,eval=FALSE}
par( mfrow=c(2,2) )
breaks = hist(c(0,zsC,zs),plot=F)$breaks
hist(zs[1:10],breaks=breaks, col=1, main='LD partners')
abline(v=zsC)
hist(zs[11:20],breaks=breaks, col=2, main='Low-LD partner SNPs')
abline(v=zsC)
hist(zs[21:30],breaks=breaks, col=3, main='Independent SNPs')
abline(v=zsC)
```
<br>

## Quite stuck here (14/10)

### __BONUS: Perform LD score regression. First, calculate the LD scores. There should be M+1 of them.__
```{r, eval=FALSE}
# sum the squared Pearson correlation coefficients (r²) of a given SNP with all other SNPs 

cor_mat = apply(gs, MARGIN = 2, function(geno) cor(geno, gs))
colSums(cor_mat^2 - 1)

# ldscores = ? .. don't know where M+1 is coming from 
ldscores = colSums(cor_mat)
ldscores 
```
<br>

### BONUS: Visualize LD score regression.
```{r,eval=FALSE}
#chis = c( ?, ? )^2
chis = c( 1, 2 )^2
plot(ldscores[1:2], chis, ylab=expression(chi^2) )
#test for inflation
lambdaGC = median(chis)/0.454
lambdaGC
```
<br>

### BONUS: Estimate heritability.
```{r,eval=FALSE}
summary( lm( ? )$coef[2,1] * M/N
```
<br>

BONUS: What is the true heritability?
```{r, eval=FALSE}
var(?) / var(?)
```









